A Dearth of Young Stars and Planets in the Kepler Field

### 2024-06-03 paper/ms.tex
update ms

Sub-gigayear planets hold particular weight in studies of planet
evolution.
While our catalog has \nplyounggyro\ confirmed planets
with median ages below 1\,Gyr, requiring $t_{\rm gyro}$$<$1\,Gyr at
2$\sigma$ confidence yields \nplyounggyrotwosigma\ planets orbiting
\nplhostsyounggyrotwosigma\ stars.  
The most notable of these planets can be split into two classes: those
in clusters, and those with independent and consistent rotation and
lithium-based ages.

@@

Four stellar ensembles in the Kepler field have to date yielded a
total of fourteen transiting planets.  Our analysis blindly recovers
the youth of these planets.

{\it Cep-Her}---The $\approx$40\,Myr Cep-Her complex has been
dated using the isochrone loci of its
pre-main-sequence M dwarfs; it contains Kepler-1627Ab, Kepler-1643b,
Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022b,Bouma_2022a}.
Different associations in the complex vary in age by $\approx$50\%
(Kerr~et al.,~submitted).  The planets
themselves are all on close-in orbits (5-25\,days), with sizes from
2-4\,$R_\oplus$.  The rotation, lithium, and cluster ages all agree
for Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
above the cluster age reported in \citet{Bouma_2022b} for RSG-5.

{\it MELANGE-3}---\citet{Barber_2022} reported a
$105\pm10$\,Myr association in the Kepler field and derived its age
using rotation, lithium, and the isochronal pre-main-sequence.
Kepler-970 and Kepler-1928 harbor two transiting mini-Neptunes inside
this association.  We find rotation-based ages
of $t_{\rm
gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
Kepler-970 and Kepler-1928 respectively.  The former rotation-based
age being slightly older than the suggested Pleiades age for the
assocation agrees with Figure~4 of \citet{Barber_2022}.

{\it Theia~520}---We derive rotation-based ages of $\approx$350\,Myr
for Kepler-968 and Kepler-52.  Based on spatial and kinematic
clustering, these stars were reported by \citet{2019AJ....158..122K}
to be members of an association called Theia~520.  The core of this
structure is also known as UBC-1 \citep{2018A&A...618A..59C}.  Based
on isochrone and rotation-based age-dating, Theia~520 seems to be
$\approx$230$\pm$70\,Myr old
\citet{2019AJ....158..122K,2024A&A...681A..13F}, which is within the
1$\sigma$ statistical uncertainties of our $t_{\rm gyro}$
measurements.  Ongoing work by \citet{Curtis2024} broadly finds that
the two planet-hosting stars are indeed members of this diffuse
population.  Given the cluster-level and individual-star evidence,
this makes Kepler-968 and Kepler-52 the youngest multiplanet systems
currently known from the main Kepler mission.

{\it NGC~6811}---\citet{Meibom_2013} reported the discovery of
Kepler-66 and -67b, two mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).
We report rotation-based ages for these systems of
\kepsixsixtgyro\,Myr and \kepsixseventgyro\,Myr for Kepler-66 and
-67b, respectively.  Kepler-66 ($T_{\rm eff}$$\approx$5900\,K, $P_{\rm
rot}$$\approx$10.4\,days) has a lower precision and a more asymmetric
posterior because of its slow spindown rate.  Kepler-67 ($T_{\rm
eff}$$\approx$5100\,K, $P_{\rm rot}$$\approx$10.3\,days) is marginally
hotter than the later K dwarfs that are in ``stalled'' spindown at
this time, enabling its rotation period to be diagnostic of its age.

@@

We compared the independent rotation and lithium ages of the
known planet hosts using the point-estimates of the available
posteriors.  There are three cases of interest:

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item Two-sided posteriors for both $t_{\rm gyro}$ and $t_{\rm Li}$:
    if their values were consistent within 2$\sigma$, we judged them
    to be consistent; if they were consistent within 2-3$\sigma$, we
    suggest them to be ``maybe'' consistent.  Otherwise, they
    disagree.
    %
  \item If $t_{\rm Li}$ is a lower limit, we compare this lower limit
    with the $1\sigma$ upper limit from $t_{\rm gyro}$; if the two
    overlapped, we judge the age estimates to be consistent.
    %
  \item If $t_{\rm Li}$ provides a two-sided posterior, and no
    rotation is reported, we judge the age estimates to be
    inconsistent.  This is because this lithium constraint can only be
    provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
    have been sensitive to the rotation periods of such stars.
\end{enumerate}

@@

Classifying Kepler stars as thin vs.~thick disk members helps connect them to 
previous work on the star formation history of the
Galaxy.  On extragalactic scales, the star formation rate in spiral
galaxies with mass similar to the Milky Way peaked $\approx$10\,Gyr
ago, and has since decreased by an order of magnitude
\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  In our own
galaxy, many investigators have worked to measure similar star
formation histories through CMD fitting of resolved stellar
populations
\citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X}
and by modeling the local white dwarf luminosity function
\citep[e.g.][]{2019ApJ...878L..11I}.  \citet{2019A&A...624L...1M} and
\citet{2019ApJ...878L..11I} for instance focused on stars (and white
dwarfs) within a few hundred parsecs, and both found a peak in the
local star formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R}
focused on stars in a wider 2\,kpc bubble, and reported three local
maxima in the thin disk's star formation history 5.7, 1.9, and
1.0\,Gyr ago, which they associated with pericenter passages of the
Sagittarius satellite galaxy.

The right column of Figure~\ref{fig:hist_tgyro} compares our derived
age distribution against star formation histories (SFHs) reported by
\citet{2019A&A...624L...1M} and \citet{2020NatAs...4..965R} based on
CMD fitting.  The overall slope of both SFHs broadly agrees with what
we find from rotation-based ages.  The \citet{2019A&A...624L...1M}
data in particular are consistent, particularly after accounting for the
statistical uncertainties reported in that study.  Regarding the
episodic star formation bursts reported by
\citet{2020NatAs...4..965R}, they are not apparent in our overall FGK
star sample (top row).  However, rotation-based ages are most precise
for $\approx$G8V-K4V dwarfs older than $\approx$1.5\,Gyr
\citep{Bouma_2023}.  The lower row of Figure~\ref{fig:hist_tgyro}
selects these stars with a temperature cut, and may show a hint of the
$\approx$2\,Gyr spike reported by \citet{2020NatAs...4..965R}.  The
$\approx$1\,Gyr spike however is not recovered.

% compare against Berger+20, note their sentence
% "Encouragingly, the distribution also qualita- tively matches the red giant asteroseismology-derived age distributions in Silva Aguirre et al. (2018) and Pinsonneault et al. (2018), as well as the rotation-based ages in Claytor et al. (2020) and the Galactic Archaeology with HERMES–Gaia ages in Buder et al. (2019)."
%
% also compare against Mor2019 and their literature assessment
% "Our findings that the thin disc SFH does not follow a sim- ple decreasing shape until the present are in good agreement with Snaith et al. (2015), and Haywood et al. (2016, 2018) who found, using data with metallicities and assuming a fixed IMF, the existence of an SFR quenching followed by a reactivation. Kroupa (2002a), using stellar kinematics, found the SFH to behave similarly. The relative maximum of the SFR that we find at 2–3 Gyr ago is compatible with the results of Vergely et al. (2002) and Cignoni et al. (2006) that, using Hipparcos data in a sphere of 80 pc around the Sun and assuming a fixed IMF, found maximum peaks at 1.75–2 Gyr ago and 2–3 Gyr ago, respec- tively. Recently, Bernard (2018), in a tentative work using TGAS data, pointed towards the existence of a relative maximum also located 2–3 Gyr ago."

%\subsection{Isochrone Age Comparison}

%\subsection{Lithium Age Comparison}

@@

\startlongtable
\begin{deluxetable*}{lllllllrrrrrr}
  \tabletypesize{\scriptsize}
  \tablecaption{
  Ages of Kepler target stars derived from rotation periods.  The
  full machine-readable table includes \nuniqstarsantosrot\ stars
  with reported rotation periods from \citetalias{Santos_2019} and
  \citetalias{Santos_2021}, \nuniqstarfinitegyroage\ of which have
  finite reported gyrochrone ages.  A random set of these stars is
  shown here to give an idea for form and content.  The quality
  bitmask is as in Table~\ref{tab:planets}; requiring bits zero to
  nine to be null yields \nuniqstarsantosrotgyroappl\ stars for
  which gyrochronology is likely to be valid.  \label{tab:stars}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KIC & Gaia DR3 &  $T_{\rm eff}$ & $P_{\rm rot}$ & $t_{\rm gyro}$ & $Q_{\rm star}$ \\
  -- &   -- & K & days &  Myr &    int  \\
  \hline
  \input{table_star_gyro} %\\
  %\hline
  \enddata
% \tablecomments{This table includes X, Y, Z... The machine-readable version,
%   available online, includes additional columns for .... 
% }
\end{deluxetable*}

>>>>

Sub-gigayear planets can be particularly informative for studies of planet
evolution.
Our catalog has \nplyounggyro\ confirmed planets
with median ages below 1\,Gyr.  Requiring $t_{\rm gyro}$$<$1\,Gyr at
2$\sigma$ yields \nplyounggyrotwosigma\ planets orbiting
\nplhostsyounggyrotwosigma\ stars.  
The youth claim is most secure for the subsets that are either
in clusters, or that have independent rotation and
lithium-based ages.

@@

Four stellar ensembles in the Kepler field have to date yielded a
total of fourteen transiting planets.  Our analysis blindly recovers
the youth of all of these planets.

{\it Cep-Her}---The Cep-Her complex ($\approx$40\,Myr)
contains Kepler-1627Ab, Kepler-1643b,
Kepler-1974b and Kepler-1975\,Ab \citep{Bouma_2022b,Bouma_2022a}.
Different sub-groups of the complex vary in age by $\approx$50\%
(Kerr~et al.,~submitted).  The planets
themselves are all on close-in orbits (5-25\,days), with sizes from
2-4\,$R_\oplus$.  Our rotation and lithium ages agree with the cluster age
for Kepler-1627A, Kepler-1974, and Kepler-1975A.  For Kepler-1643, the
rotation and cluster ages agree, and the lithium age is 2.0$\sigma$
above the cluster age reported in \citet{Bouma_2022b} for RSG-5.

{\it MELANGE-3}---\citet{Barber_2022} reported a
$105\pm10$\,Myr association in the Kepler field
containing two transiting mini-Neptunes:
Kepler-970 and Kepler-1928.  We find rotation-based ages
of $t_{\rm
gyro}=176_{-40}^{+120}$\,Myr and $144_{-88}^{+104}$\,Myr for
Kepler-970 and Kepler-1928 respectively.  The former rotation-based
age being slightly older than the suggested Pleiades age for the
association agrees with Figure~4 of \citet{Barber_2022}.

{\it Theia~520}---We derive rotation-based ages of $\approx$350\,Myr
for Kepler-968 and Kepler-52.  Based on spatial and kinematic
clustering, these stars were reported by \citet{2019AJ....158..122K}
to be members of Theia~520.  The core of this
structure is also known as UBC-1 \citep{2018A&A...618A..59C}.  Based
on isochrone and rotation-based age-dating, Theia~520 seems to be
$\approx$230$\pm$70\,Myr old
\citep{2019AJ....158..122K,2024A&A...681A..13F}, which is within 
1$\sigma$ of our $t_{\rm gyro}$
measurements.  Ongoing work by \citet{Curtis2024} broadly finds that
the two planet-hosting stars are indeed members of this diffuse
population.  Given the cluster-level and individual-star evidence,
this makes Kepler-968 and Kepler-52 the youngest multiplanet systems
currently known from the main Kepler mission.

{\it NGC~6811}---\citet{Meibom_2013} reported the discovery of
Kepler-66 and -67b, two mini-Neptunes in NGC~6811 ($\approx$1\,Gyr).
We derive rotation-based ages for these systems of
\kepsixsixtgyro\,Myr and \kepsixseventgyro\,Myr for Kepler-66 and
-67b, respectively.  Kepler-66 ($T_{\rm eff}$$\approx$5900\,K, $P_{\rm
rot}$$\approx$10.4\,days) has a lower precision and a more asymmetric
posterior because of its slow spindown rate.  Kepler-67 ($T_{\rm
eff}$$\approx$5100\,K, $P_{\rm rot}$$\approx$10.3\,days) is marginally
hotter than the later K dwarfs that are in ``stalled'' spindown at
this time, enabling its rotation period to be diagnostic of its age.

@@

The comment column of Table~\ref{tab:planets} compares
the independent rotation and lithium ages of the
known planet hosts using the point-estimates of their age posteriors.
There are three cases of interest.
{\it 1i)} If two-sided posteriors for both $t_{\rm gyro}$ and $t_{\rm Li}$ existed,
and their values were consistent within 2$\sigma$, we judged them
to be consistent; if they were consistent within 2-3$\sigma$, we
listed them as ``maybe'' consistent.  Otherwise, they
disagreed.
{\it ii)}
If $t_{\rm Li}$ was a lower limit, we compared this lower limit
with the $1\sigma$ upper limit from $t_{\rm gyro}$; if the two
overlapped, we judged the age estimates to be consistent.
{\it iii)}
If $t_{\rm Li}$ provided a two-sided posterior, and no
rotation period was found, we judged the age estimates to be
inconsistent.  This is because a two-sided lithium constraint can only be
provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
have been sensitive to the rotation periods of such stars.

%\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
%  \item Two-sided posteriors for both $t_{\rm gyro}$ and $t_{\rm Li}$:
%    if their values were consistent within 2$\sigma$, we judged them
%    to be consistent; if they were consistent within 2-3$\sigma$, we
%    suggest them to be ``maybe'' consistent.  Otherwise, they
%    disagree.
%    %
%  \item If $t_{\rm Li}$ is a lower limit, we compare this lower limit
%    with the $1\sigma$ upper limit from $t_{\rm gyro}$; if the two
%    overlapped, we judge the age estimates to be consistent.
%    %
%  \item If $t_{\rm Li}$ provides a two-sided posterior, and no
%    rotation is reported, we judge the age estimates to be
%    inconsistent.  This is because this lithium constraint can only be
%    provided for G and K stars $\lesssim$2\,Gyr old, and Kepler should
%    have been sensitive to the rotation periods of such stars.
%\end{enumerate}

@@

Classifying Kepler stars as thin vs.\ thick disk members helps connect
them to previous work on the star formation history of the Galaxy.  On
extragalactic scales, the star formation rate in spiral galaxies with
mass similar to the Milky Way peaked $\approx$10\,Gyr ago, and has
since decreased by an order of magnitude
\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  In our own
galaxy, many investigators have worked to measure similar star
formation histories through CMD fitting of resolved stellar
populations
\citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X}
and by modeling the local white dwarf luminosity function
\citep[e.g.][]{2019ApJ...878L..11I}.  \citet{2019A&A...624L...1M} and
\citet{2019ApJ...878L..11I} for instance focused on stars (and white
dwarfs) within a few hundred parsecs, and both found a peak in the
local star formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R}
focused on stars in a wider 2\,kpc bubble, and reported three local
maxima in the thin disk's star formation history 5.7, 1.9, and
1.0\,Gyr ago, which they associated with pericenter passages of the
Sagittarius satellite galaxy.

The right column of Figure~\ref{fig:hist_tgyro} compares our derived
age distribution against star formation histories (SFHs) reported by
\citet{2019A&A...624L...1M} and \citet{2020NatAs...4..965R} based on
CMD fitting.  The overall slope of both SFHs broadly agrees with what
we find from rotation-based ages.  The \citet{2019A&A...624L...1M}
data in particular are consistent, particularly after accounting for
the statistical uncertainties reported in that study.  Regarding the
episodic star formation bursts reported by
\citet{2020NatAs...4..965R}, they are not apparent in our overall FGK
star sample (top row).  However, rotation-based ages are most precise
for $\approx$G8V-K4V dwarfs older than $\approx$1.5\,Gyr
\citep{Bouma_2023}.  The lower row of Figure~\ref{fig:hist_tgyro}
selects these stars with a temperature cut, and may show a hint of the
$\approx$2\,Gyr spike reported by \citet{2020NatAs...4..965R}.  The
$\approx$1\,Gyr spike however is not recovered.  This
$\approx$2.5\,Gyr local maximum is similarly present in the isochrone
ages derived by \citet{Berger_2020a_catalog} for the Kepler field, and
in the red giant asteroseismic ages derived by
\citet{SilvaAguirre2018}.  The main novelty of our ages relative to
those studies is their accuracy at $<$1\,Gyr.

@@

\startlongtable
\begin{deluxetable*}{lllllllrrrrrr}
  \tabletypesize{\scriptsize}
  \tablecaption{
  Ages of Kepler target stars derived from rotation periods.  The
  full machine-readable table includes \nuniqstarsantosrot\ stars
  with reported rotation periods from \citetalias{Santos_2019} and
  \citetalias{Santos_2021}, \nuniqstarfinitegyroage\ of which have
  finite reported gyrochrone ages.  A random set of these stars is
  shown to give an idea for form and content.  The quality
  bitmask is as in Table~\ref{tab:planets}; requiring bits zero to
  nine to be null yields \nuniqstarsantosrotgyroappl\ stars for
  which gyrochronology is likely to be valid.  \label{tab:stars}}
  %\toprule
  %\midrule
  %\endhead
  \startdata
  KIC & Gaia DR3 &  $T_{\rm eff}$ & $P_{\rm rot}$ & $t_{\rm gyro}$ & $Q_{\rm star}$ \\
  -- &   -- & K & days &  Myr &    int  \\
  \hline
  \input{table_star_gyro} %\\
  %\hline
  \enddata
% \tablecomments{This table includes X, Y, Z... The machine-readable version,
%   available online, includes additional columns for .... 
% }
\end{deluxetable*}
